Site-Heterogeneous Character Change Models for Morphology

### 2019-01-03 Papers/DraftOutlineSSCCM.tex
adding results desc of barden tree, fig

Morphological data is challenging to model due to the complexity of morphological characters relative to nucleotide sequence or amino acid characters. 
The Mk model, a generalization of the Jukes-Cantor model, assumes that there is an equal rate of forward and backwards transitions in morphological traits.
In this study, we use an expanded combined set of two ant (Formicidae) data matrices, composed of one matrix of extinct and extant ants and one matrix of extant ants to estimate a phylogeny for the group.
We use this large matrix to explore methods for relaxing the transition rate symmetry of the Mk model, and introduce the Site Heterogeneous Discrete Morphology (SHDM) to allow equilibrium character frequencies to vary. 
These two models can be used to explicitly model different character frequencies within one dataset, as opposed to assuming a single common mechanism with respect to among character frequency variation.

@@

	Recent interest in using Bayesian methods to model morphological evolution to estimate phylogenetic trees has spurred many studies into how well the existing toolkit performs, particularly in relation to parsimony methods.

@@

	The most common way to perform likelihood or Bayesian phylogenetic analysis using morphological data is by applying a model of morphological evolution called the Mk model.

@@

	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two states is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.

@@

	This has complicated the ability to specify $\alpha$ parameters to the Q-matrix \textit{a priori}.
	This is a stark contrast to molecular data, in which base pairs or amino acid residues are generally assumed to have similar properties across an alignment, and exceptions are often defined with respect to known features, such as stem and loop domains. \par

@@

Likewise, a highly-probable changes may be seen relatively rarely if the starting character is observed rarely. 

@@

Even with expanded taxon sampling in the Ponerinae, there is still conflict between the molecular and morphological estimates of phylogenetic relationships. \par

@@

We also perform simulations to demonstrate that the true number of transition rate asymmetry categories is detectable from the data, and that using an appropriately-specified model of morphological evolution is less likely to lead to incorrect phylogenetic inference than a correctly-specified model.\par

@@

We do, however, support the monophyly of what  has been considered the crown group of ants, and support a monophyletic Haidomyrmex ants, as in Barden and Grimaldi (2017). 

@@

The marginal likelihood comparisons for the Barden and Grimaldi dataset can be seen in Fig. 1a.
The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 1.
In the Barden dataset, the data support 4 character state frequency variation.
In the combined dataset, 6 categories are supported for binary data and 7 are supported for multistate. \par
When a phylogeny is estimated from the Barden dataset under the supported model, it is largely unresolved.

@@

In the large datasets, underparameterized models have a deleterious impact on the accuracy of phylogenetic analyses. 
However, overparameterized models did not show a similar decrease in accuracy.
In the smaller datasets, accuracy does not seem to be tied to the parameterization of the model. \par

@@

In the large, multistate data sets, accuracy is improved using a correctly-specified model over both over- and underparameterized models. 
In the small datasets, a pattern similar to that of the large binary datasets is observed, in which underparameterization negatively effected accuracy, while overparameterization did not. \par

@@

@@

WHY WON'T YOU RENDER, THING?

@@

\bibliographystyle{sysbio.bst}
\bibliography{bibliography}

>>>>

Morphological data is challenging to model due to the complexity of morphological characters relative to nucleotide sequence or amino acid characters. 
The Mk model is the most commonly-used model for incorporating morphology in maximum likelihood and Bayesian work.
This model is a generalization of the Jukes-Cantor model, assumes that there is an equal rate of forward and backwards transitions in morphological traits.
In this study, we explore the use of models that allow asymmetrical transition rates.
We use an expanded combined set of two ant (Formicidae) data matrices, composed of one matrix of extinct and extant ants and one matrix of extant ants to estimate a phylogeny for the group.
In particular, we examine the use of two models to allow the equilibrium character frequency to vary. 
One model, a Beta prior on state frequencies is similar to a model implemented in MrBayes and can be used to allow asymmetrical transition rates in binary characters.
We also describe a Dirichlet model to allow asymmetrical transitions in multistate characters.
These two models can be used to explicitly model different character frequencies within one dataset, as opposed to assuming a single common mechanism with respect to among character frequency variation.

@@

	Recent interest in using Bayesian methods to model morphological evolution to estimate phylogenetic trees has spurred many studies into how well the existing phylogenetic models perform, particularly in relation to parsimony methods.

@@

	The most common way to perform likelihood or Bayesian phylogenetic analysis using discrete morphological data is by applying a model of morphological evolution called the Mk model.

@@

	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two or more states  is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.

@@

	This complicates researchers' ability to specify unique $\alpha$ parameters to the Q-matrix, or to apply multiple Q-matrices across the dataset.
	While the basic model is very simple, some elaborations of the model are common.
	Assumptions about character evolution, such as ordering, have been common in parsimony analysis, and can be accommodated under the Mk model. 
	This is a stark contrast to molecular data, in which base pairs or amino acid residues are generally assumed to have similar properties across an alignment, and exceptions are often defined with respect to known features, such as stem and loop domains. 

@@

Likewise, a highly probable changes may be seen relatively rarely if the starting character is observed rarely. 

@@

Even with expanded taxon sampling in the Ponerinae, there is still conflict between the molecular and morphological estimates of phylogenetic relationships. 
Absent inputs of additional information, such as clade constraints, the trees are also substantially unresolved. \par

@@

We also perform simulations to demonstrate that the true number of transition rate asymmetry categories is detectable from the data, and that using an appropriately-specified model of morphological evolution is less likely to lead to incorrect phylogenetic inference than a correctly specified model.\par

@@

We do, however, support the monophyly of what  has been considered the crown group of ants, and support a monophyletic Haidomyrmicine ants, as in Barden and Grimaldi (2017). 

@@

The marginal likelihood comparisons for the Barden and Grimaldi dataset can be seen in Fig. 2C and 2D.
The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 2A, and including the multistate characters can be seen in 2B.
In the Barden datasets, the data support 4 character state frequency variation categories for both the binary and multistate data.
In the combined dataset, 6 categories are supported for binary data and 7 are supported for multistate. \par
When a phylogeny is estimated from the Barden dataset under the supported model, it is largely unresolved.

@@

In the large datasets, underparameterized models have a deleterious impact on the accuracy of phylogenetic analyses. 
However, overparameterized models did not show a similar decrease in accuracy.
In the smaller datasets, accuracy does not seem to be tied to the parameterization of the model. 
Overall, phylogenetic error is lower in larger datasets, even with appropriately parameterized models. \par

@@

In the large, multistate data sets, accuracy is improved using a correctly-specified model over both over- and underparameterized models. 
In the small datasets, a pattern similar to that of the large binary datasets is observed, in which underparameterization negatively effected accuracy, while overparameterization did not. 
As with binary data, dataset size has a strong relationship with phylogenetic accuracy. 
However, the multistate datasets show a stronger relationship between appropriate parameterization and accuracy. 
This is especially obvious in the small multistate datasets, in which choosing a correct model improves accuracy by approximately 40\%. \par

@@

In simulation, for larger datasets, underparameterization of the morphological model appears to have more serious deleterious effects on accuracy than does overparameterization.
This is expected, as statistical literature indicates that underparameterization often leads to error, but overparameterization often leads to less certainty.
In the small datasets, we observe different dynamics. 
The binary datasets do not show a relationship between the appropriateness of the model and accuracy.
These datasets are quite small - 41 taxa and 26 characters, modeled on Barden and Grimaldi (2015). 
This dataset is on the extreme small end of the datasets explored by Wright et al. (2016). 
For the multistate data, the choice of model greatly affects accuracy. 
Under an overly-simple model of evolution, phylogenetic error is nearly 40\%. 
But under the correctly-specified model, the error is cut by about 40\%. 
Why should these two datasets show such differences? 
It's possible that there is simply not enough information to support a more complex model for the binary data.
The multistate dataset has nearly twice as many characters for the same amount of taxa. 
Secondly, with more character states, and more state frequency parameters, the model violations associated with under- and overparameterization are more severe in the multistate datasets.

@@

Ants have a rich fossil record.
Due to the presence of stem ants in this fossil record, the assumptions of the Mk model are violated in the data.
Particularly, due to the loss of wasp-like characters between stem and crown ants, and the gain of ant family and subfamily apomorphies, a single mechanism specifying symmetric transitions is a poor fit to these data. 
Here, we use a discrete Beta model to estimate phylogenetic trees from binary data, and the Site-Heterogeneous Discrete Morphology model to estimate trees from binary and multistate datasets.
We also use simulation to test these models. 
We find that using a correctly-specified model can decrease the chances of estimating an incorrect tree.
Despite these advances, the tree of ants from morphological data remains unresolved.
However, this model still enables researchers to make assumptions that may strike them as more realistic about their individual datasets.
\par

@@

\bibliographystyle{sysbio}
\bibliography{bibliography.bib}
